Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-03-31 rmd/manuscript.Rmd
edit: manuscript

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: the-american-naturalist.csl
header-includes:
  \usepackage{textgreek}
---

@@

```{=tex}
\begin{align}
logit~m_{down} = logit~m_{up}+ \epsilon
\end{align}
```
where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles a natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches):

```{=tex}
\begin{align}
logit~m_{down} = \omega~(logit~m_{1,up} +\epsilon_1)+(1-\omega)~(logit~m_{2,up} +\epsilon_2)
\end{align}
```
$m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\epsilon_1$ and $\epsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

```{=tex}
\begin{align}
K_{x} = K_{base}(N_{p,up} + 1)^z
\end{align}
```
where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on an empirical relationship of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

```{=tex}
\begin{align}
A_{i,x}(t+1) \sim Poisson(\bar{A}_{i,x}(t) + I_{i,x}(t) - E_{i,x}(t))
\end{align}
```
where $\bar{A}_{i,x}(t+1)$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

@@

```{=tex}
\begin{align}
A_{B,x}'(t) = s_{0,B}[1-z_{B,x}(t)]A_{B,x}(t)
\end{align}
```
where $s_{0,i}$ is the background survival probability (density independent) and $z_{B,x}(t)$ the proportional mortality caused by predation. Let $W_{B,x}(t)$ be the total number of basal species consumed by $C$ and $P$. The predator-induced mortality $z_{B,x}(t)$, the sum of proportional predation by $C$ ($z_{BC,x}(t)$) and $P$ ($z_{BP,x}(t)$), is then:

```{=tex}
\begin{equation}
z_{B,x}(t) = z_{BC,x}(t) + z_{BP,x}(t) =
\left\{
\begin{aligned}
&\frac{W_{B,x}(t)}{A_{B,x}(t)} &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&1 &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
and

```{=tex}
\begin{align}
W_{B,x}(t) = W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)
\end{align}
```
where $W_{ij,x}(t)$ is the population-wide consumption of consumer $j$ on prey $i$ (number of individuals), and the parameter $\delta_{x}(t)$ is the relative preference of predator to basal species over consumer. We assume the Holling Type-II functional response for predator-prey interactions:

```{=tex}
\begin{align}
W_{ij,x}(t) = \frac{a_{ij}A_{i,x}(t)A_{j,x}(t)}{A_{j,x}(t) + a_{ij}h_{ij}A_{i,x}(t)}
\end{align}
```
$a_{ij}$ is the attack rate and $h_{ij}$ the handling time. To be realistic, the preference parameter is assumed to depend on the relative potential amount of energy obtained from each prey:

```{=tex}
\begin{align}
\delta_{x}(t) = \frac{e_{BP} f_{BP}(A_{B,x}(t))}{e_{BP} f_{BP}(A_{B,x}(t)) + e_{CP} f_{CP}(A_{C,x}(t))}
\end{align}
```
where $e_{ij}$ is the conversion efficiency from prey $i$ to consumer $j$, and $f_{ij}(A_{i,x}(t)) = \frac{a_{ij}A_{i,x}(t)}{1 + a_{ij}h_{ij}A_{i,x}(t)}$. An analogous approach has been applied to a model with multiple prey choices [@nishijimaRolesAlternativePrey2014]. Subsequently, a density-dependent recruitment occurs:

```{=tex}
\begin{align}
\hat{A}_{B,x}(t) = \frac{r_{b} A_{B,x}'(t)}{1 + \frac{r_b - 1}{K_{x}}A_{B,x}'(t)}
\end{align}
```
$r_b$ is the reproductive rate for survived individuals.

@@

```{=tex}
\begin{equation}
\hat{A}_{C,x}(t) =
\left\{
\begin{aligned}
&s_{0,C} (1 - z_{CP,x}) e_{BC}W_{BC,x}(t) &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&s_{0,C} (1 - z_{CP,x}) e_{BC} \left( \frac{W_{BC,x}(t)}{W_{B,x}(t)} A_{B,x} \right) &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
Equation 11 assumes competition between $C$ and $P$ when their total consumption $W_{B,x}(t)$ exceeds the abundance of basal species $A_{B,x}(t)$, and the relative consumption $\frac{W_{BC,x}(t)}{W_{B,x}(t)}$ rests on $a_{ij}$, $h_{ij}$, and $A_{j,x}(t)$.

@@

```{=tex}
\begin{equation}
\hat{A}_{P,x}(t) =
\left\{
\begin{aligned}
&s_{0,P} \left[e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&s_{0,P} \left[e_{BP} \left(\frac{\delta_{x}(t)W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
After recruitment, the constituent species experience stochastic regional disturbance, which causes proportional loss of $m$ to populations:

```{=tex}
\begin{align}
\bar{A}_{i,x}(t) = [1 - \psi(t) m_{x}] \hat{A}_{i,x}(t)
\end{align}
```
The binary random variable $\psi(t)$ expresses the occurrence of regional disturbance (disturbance occurs if $\psi(t)=1$ otherwise $0$), and $m_x$ is the disturbance intensity at patch $x$ as determined by the network generation process. $\psi(t)$ is drawn from a Bernoulli distribution with disturbance probability $p_m$.

@@

```{=tex}
\begin{align}
\xi_{i,x}(t)=\frac{\sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})}{\sum_{x} \sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})}
\end{align}
```
where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{i,x}(t)=\xi_{i,x}(t) \sum_{x}E_{i,x}(t)$.

@@

```{=tex}
\begin{align}
TP_{P,x}(t) = \delta_{x}(t) + 2(1 - \delta_{x}(t)) + 1
\end{align}
```
This definition accounts for the influence of omnivory.

>>>>

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  #pdf_document:
  #  latex_engine: xelatex
  word_document:
    reference_docx: word_format.docx
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: the-american-naturalist.csl
header-includes:
  \usepackage{textgreek}
---

@@

$$
logit~m_{down} = logit~m_{up}+ \epsilon
$$

where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles a natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches):

$$
logit~m_{down} = \omega~(logit~m_{1,up} +\epsilon_1)+(1-\omega)~(logit~m_{2,up} +\epsilon_2)
$$

$m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\epsilon_1$ and $\epsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

$$
K_{x} = K_{base}(N_{p,up} + 1)^z
$$

where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on an empirical relationship of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

$$
A_{i,x}(t+1) \sim Poisson(\bar{A}_{i,x}(t) + I_{i,x}(t) - E_{i,x}(t))
$$

where $\bar{A}_{i,x}(t+1)$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

@@

$$
A_{B,x}'(t) = s_{0,B}[1-z_{B,x}(t)]A_{B,x}(t)
$$

where $s_{0,i}$ is the background survival probability (density independent) and $z_{B,x}(t)$ the proportional mortality caused by predation. Let $W_{B,x}(t)$ be the total number of basal species consumed by $C$ and $P$. The predator-induced mortality $z_{B,x}(t)$, the sum of proportional predation by $C$ ($z_{BC,x}(t)$) and $P$ ($z_{BP,x}(t)$), is then:

$$
z_{B,x}(t) = z_{BC,x}(t) + z_{BP,x}(t) =\left\{\begin{aligned}&\frac{W_{B,x}(t)}{A_{B,x}(t)} &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&1 &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

and

$$
W_{B,x}(t) = W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)
$$

where $W_{ij,x}(t)$ is the population-wide consumption of consumer $j$ on prey $i$ (number of individuals), and the parameter $\delta_{x}(t)$ is the relative preference of predator to basal species over consumer. We assume the Holling Type-II functional response for predator-prey interactions:

$$
W_{ij,x}(t) = \frac{a_{ij}A_{i,x}(t)A_{j,x}(t)}{A_{j,x}(t) + a_{ij}h_{ij}A_{i,x}(t)}
$$

$a_{ij}$ is the attack rate and $h_{ij}$ the handling time. To be realistic, the preference parameter is assumed to depend on the relative potential amount of energy obtained from each prey:

$$
\delta_{x}(t) = \frac{e_{BP} f_{BP}(A_{B,x}(t))}{e_{BP} f_{BP}(A_{B,x}(t)) + e_{CP} f_{CP}(A_{C,x}(t))}
$$

where $e_{ij}$ is the conversion efficiency from prey $i$ to consumer $j$, and $f_{ij}(A_{i,x}(t)) = \frac{a_{ij}A_{i,x}(t)}{1 + a_{ij}h_{ij}A_{i,x}(t)}$. An analogous approach has been applied to a model with multiple prey choices [@nishijimaRolesAlternativePrey2014]. Subsequently, a density-dependent recruitment occurs:

$$
\hat{A}_{B,x}(t) = \frac{r_{b} A_{B,x}'(t)}{1 + \frac{r_b - 1}{K_{x}}A_{B,x}'(t)}
$$

$r_b$ is the reproductive rate for survived individuals.

@@

$$
\hat{A}_{C,x}(t) =\left\{\begin{aligned}&s_{0,C} (1 - z_{CP,x}) e_{BC}W_{BC,x}(t) &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,C} (1 - z_{CP,x}) e_{BC} \left( \frac{W_{BC,x}(t)}{W_{B,x}(t)} A_{B,x} \right) &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

The above equation assumes competition between $C$ and $P$ when their total consumption $W_{B,x}(t)$ exceeds the abundance of basal species $A_{B,x}(t)$, and the relative consumption $\frac{W_{BC,x}(t)}{W_{B,x}(t)}$ rests on $a_{ij}$, $h_{ij}$, and $A_{j,x}(t)$.

@@

$$
\hat{A}_{P,x}(t) =\left\{\begin{aligned}&s_{0,P} \left[e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,P} \left[e_{BP} \left(\frac{\delta_{x}(t)W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

After recruitment, the constituent species experience stochastic regional disturbance, which causes proportional loss of $m$ to populations:

$$
\bar{A}_{i,x}(t) = [1 - \psi(t) m_{x}] \hat{A}_{i,x}(t)
$$

The binary random variable $\psi(t)$ expresses the occurrence of regional disturbance (disturbance occurs if $\psi(t)=1$ otherwise $0$), and $m_x$ is the disturbance intensity at patch $x$ as determined by the network generation process. $\psi(t)$ is drawn from a Bernoulli distribution with disturbance probability $p_m$.

@@

$$
\xi_{i,x}(t)=\frac{\sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})}{\sum_{x} \sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})}
$$

where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{i,x}(t)=\xi_{i,x}(t) \sum_{x}E_{i,x}(t)$.

@@

$$
TP_{P,x}(t) = \delta_{x}(t) + 2(1 - \delta_{x}(t)) + 1
$$

This definition accounts for the influence of omnivory.
